---
title: "Ashish Bhandari Project 1"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Project 

Develop a set of Logistic regression model from the 89 specimens that you can use to predict the group membership of the remaining 199 specimens’. 

1. Format the data contained in the excel spreadsheet for use in R.

2. Perform an exploratory data analysis.

3.	Explain your GLM and assess the quality of the fit with the classified observations.   
  -	Use Cross Validation to predict the accuracy of your model.

4.	Provide a one-page write-up (excluding graphs, tables and figures) explaining your analysis of the dataset and your recommendations on the usefulness of your predictions. 

5.	Provide predictions for the unclassified observations. 		
6.	As a secondary component provide annotated code that replicates your analysis.
	

### Source
Airoldi, J.-P., B. Flury, M. Salvioni (1996) "Discrimination between two species of Microtus using both classified and unclassified observations" Journal of Theoretical Biology 177:247-262

## Vole Skull Excell Spread Sheet 

### Description
Microtus multiplex and M. subterraneus are difficult to distinguish morphologically. Here we have 43 known multiplex voles, 46 known subterraneus voles and a further 199 unidentified voles.

#### Data
288 Skulls from 2 species of voles with up to 3 measurements on  each of the vole skulls.  Note that some of the skulls labeled as unknown are missing some of the features. So that you will have to build separate models for the skulls that only have a subset of the features.

#### Group
a factor with levels 
-multiplex 
-subterraneus 
-unknown

#### Length
Condylo incisive length or skull length (0.01mm)

#### Height
Skull height above bullae (0.01mm)

#### Rostrum
Skull width across rostrum (0.01mm)

### Details
89 species have been identified by means of chromosomal analysis.

### Overview:
For this assigned project, we are given datasets on two different species of 
Microtus. Microtus is a genus of voles found in North America, Europe and 
northern Asia. From the given datasets of two different species *multiplex* 
and *subterraneus*, there are a total 43 known *multiplex* voles, 46 
known *subterraneus* voles and a total of 199 unidentified voles.

Firstly, we load different libraries that is required for this project.
```{r libraries, message=FALSE}
# loading libraries
#install.packages("GGally")
#install.packages("rpart")
library(readxl)
library(tidyverse)
library(boot)
library(dplyr)
library(ggplot2)
library(GGally)
library(rpart)
```
We will be extracting datasets from our folder and assigning them, renaming
them with same column names for easiness and then combining our datasets.
```{r data}
# loading data from the excel sheet
data3 <- read_excel("C:/Users/ashis/Documents/Dakota State University/STAT 601/Project 1 new/Vole Skulls.xlsm", sheet = "Unknown")
data2 <- read_excel("C:/Users/ashis/Documents/Dakota State University/STAT 601/Project 1 new/Vole Skulls.xlsm", sheet = "Subterraneus")
data1 <- read_excel("C:/Users/ashis/Documents/Dakota State University/STAT 601/Project 1 new/Vole Skulls.xlsm", sheet = "Multiplex")


# renaming column names of each datasets
names(data1) <- c("Index", "Details", "Length", "Height", "Width")
names(data2) <- c("Index", "Details", "Length", "Height", "Width")
names(data3) <- c("Index", "Details", "Length", "Height", "Width")

# combining all 3 datasets
new_data <- rbind(data1, data2)
```
Some outliers are due to the natural variations in the population, and they
should be left as is in our datasets. These are called true outliers. Other 
outliers are impossible values to obtain and should be removed because they 
represent measurement errors, data entry or processing errors, or poor sampling.
In case of the dataset provided to us, there were some datasets that looked
like data entry error. Those entries (rows) were replaced with NA values and 
later were removed from the datasets.
```{r}
# finding out outliers in the datasets

length_out1 <- boxplot.stats(new_data$Length)
length_out1$out

length_out2 <- boxplot.stats(data3$Length)
length_out2$out

height_out1 <- boxplot.stats(new_data$Height)
height_out1$out

height_out2 <- boxplot.stats(data3$Height)
height_out2$out

weight_out1 <- boxplot.stats(new_data$Width)
weight_out1$out

weight_out2 <- boxplot.stats(data3$Width)
weight_out2$out

# Replacing outliers with NA values

new_data[new_data$Length %in% length_out1$out, "Length"] = NA
new_data[new_data$Height %in% height_out1$out, "Height"] = NA
new_data[new_data$Width %in% weight_out1$out, "Width"] = NA

data3[data3$Length %in% length_out2$out, "Length"] = NA
data3[data3$Height %in% height_out2$out, "Height"] = NA
data3[data3$Width %in% weight_out2$out, "Width"] = NA

# removing the rows that have N/As
new_voles_data <- new_data[complete.cases(new_data), ]
summary(new_voles_data)

data_unknown <- data3[complete.cases(data3), ]

plot_data1 <- plot(new_voles_data$`Height`~ new_voles_data$`Length`,  
                   main = "Height vs Length", xlab = "Height of the skull",
                   ylab = "Length of the skull")

plot_data2 <- plot(new_voles_data$`Width` ~ new_voles_data$`Length`, 
                   main = "Width vs Length", xlab= "Width of the skull",
                   ylab = "Length of the skull")

plot_data3 <- plot(new_voles_data$`Height`~ new_voles_data$`Width`,
                   main = "Height vs Width", xlab = "Height of the skull", 
                   ylab = "Width of the skull")

# boxplots using different variables
length_plot <- boxplot(new_voles_data$Length, main = "Skull length of known 
vole species", xlab = "Skull length", ylab = "Length in mm", col = "green")
length_plot

height_plot <- boxplot(new_voles_data$Height, main = "Skull height of known vole 
                       species",xlab = "Skull height", ylab = "Height in mm",
                       col = "red")
height_plot

width_plot <- boxplot(new_voles_data$Width, main = "Skull width of known vole
            species", xlab = "Skull width", ylab = "Width in mm", col = "blue")
width_plot

```
From our scatterplots, we can derive that there is a linear relationship
between height vs length or length vs width or height vs width.

```{r}
# We have to use ifelse statement to assign 1s and 0s for 
# *subterraneus* and 0s for *multiplex* respectively.

new_voles_data$Index <- ifelse(new_voles_data$Details == "multiplex", 0,1)
new_voles_data

# finding out proportions for the dataset
propn <- prop.table(table(new_voles_data$Details))
propn
```
The proportion values for *multiplex* species and *subterraneus* are: 0.4878049 
and 0.5121951 respectively. This indicates they are almost as same proportions 
in the dataset.
```{r Plot Correlatiopn, message=FALSE}
# co-relation plotting
plot_corr <- ggpairs(data = new_voles_data, )
plot_corr
```
We find out that the correlation between width and length is stronger. It has a
positive co-relation of 0.873.
```{r}
# Density plots
ggplot_length <- ggplot(new_voles_data, aes(Length)) + geom_density() + 
  labs(x="Skull Length", y="Density", title = "Skull Length")
ggplot_length

ggplot_width <- ggplot(new_voles_data, aes(Width)) + geom_density() + 
  labs(x="Skull Length", y="Density", title = "Skull Width")
print(ggplot_width)

ggplot_height <- ggplot(new_voles_data, aes(Height)) + geom_density() +
  labs(x="Skull Length", y="Density", title = "Skull Height")
ggplot_height

set.seed(50)
# dividing the dataset to 80/20 split for training and testing
sample_data <- sample(1:nrow(new_voles_data),0.80*nrow(new_voles_data))

train_data <- new_voles_data[(sample_data), ] 
test_data <- new_voles_data[-sample_data, ]

# Now, we build our models to find out the best amongst them

model1 <- glm(as.factor(Details) ~ Length, data = train_data, 
              family = binomial())
summary(model1)

model2 <- glm(as.factor(Details) ~ Height, data = train_data, 
              family = binomial())
summary(model2)

model3 <- glm(as.factor(Details) ~ Width, data = train_data,
              family = binomial())
summary(model3)

model4 <- glm(as.factor(Details)~ Length + Height, data = train_data, 
              family = binomial())
summary(model4)

model5 <- glm(as.factor(Details)~ Length + Width, data = train_data,
              family = binomial())
summary(model5)

model6 <- glm(as.factor(Details)~ Length + Height + Width, 
              data = train_data, family = binomial())
summary(model6)
```
One of the ways to find out the effectiveness of a model is Cross-validation
technique. We have used K-fold cross validation for our models.
```{r}

# cross validation errors for our model
error1 <- cv.glm(train_data, model1, K = 65)
error1$delta[1]

error2 <- cv.glm(train_data, model2, K = 65)
error2$delta[1]

error3 <- cv.glm(train_data, model3, K = 65)
error3$delta[1]

error4 <- cv.glm(train_data, model4, K = 65)
error4$delta[1]

error5 <- cv.glm(train_data, model5, K = 65)
error5$delta[1]

error6 <- cv.glm(train_data, model6, K = 65)
error6$delta[1]

```
In the above cross validation calculations, we have found out that the
error for model 6 is 0.096, which is far lesser compared to other models.
Thus, we can say that model 6 is the best model for future predictions.
```{r}

#predicting the unknown species in the dataset
pred_uknown <- rpart(as.factor(Details)~ Length + Height + Width,
                     data = train_data)
predict(pred_uknown, newdata = test_data, type="class")
```
Finally, we were able to predict the results of our unknown species in the
given datasets. Our model had a error rate of 9.6% which is great for our 82
known species.